---
title: "Mediation"
author: "Patrick Wilson MPH"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
solarized_dark(
  code_font_family = "Fira Code",
  code_font_url    = "https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css"
)
```


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE)

library(arsenal)
library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
```

## What we hope to go over

--

- Background 

--

- Causal Estimands and DAGS

--

- Mediation Assessment and Heuristics


--

- Example of Mediation Analyses

--

- Resources

---
class: inverse, middle, center
background-size: contain

# What is mediation analyses

> _"The presence or absence of statistical interaction between two factors depends on the scale with which one chooses to measure their association" -Modern Epidemiology_


---
## Overview 

Basic defintion of Mediation analyses. Usually constrated with interaction analyses. The how vs for whom quote

"How does the cause affect the outcome?"
"For whom does and effect occur?"

Well suited for understanding mechanisms 


---
Formalized in Baron & Kenny with linear regression (1986), 

simple model known as the product method. 


---
## What does Causal inference add?

Inherently a causal question 
Baron and Kenny approach limitations with respect to interactions and nonlinearitiies

PO framework great to integrate these ideas along with proper confounding control 

Greenland Robins and Pearl subsequently formulated Causal mediation analyses and developed the causal notion of Direct effect and Indirect effect.


---
## What kind of questions can you ask?

- Pain Example

- Depression Example

- Biomarker Example

- Reproductive Epidemiology Example

- Social Epidemiology Example



- Let us use a classic example in observational epidemiology to illustrate the issue of scale

- Here you have a hypothetical cohort looking at the interaction between smoking and asbestos on the effect of getting lung cancer per 1,000 persons.


```{r, echo = FALSE}
data_2b2 <- data.frame( status = c("Non-smoker", "Smoker"), `No Asbestos` = c("$p_{00}$ = 1", "$p_{01}$ = 10"), `Asbestos` = c("$p_{10}$ = 3", "$p_{11}$ = 20")) %>% 
  rename(` ` = status)

kable(data_2b2, "html", escape =F) %>% 
  kable_styling("striped", full_width = F, font_size = 20) 
```

- With these four risks you can calculate everything you need to assess interaction.



---
class: inverse, middle, center
background-size: contain

# What is mediation analyses

> _"The presence or absence of statistical interaction between two factors depends on the scale with which one chooses to measure their association" -Modern Epidemiology_


---
## Causal Estimands and DAGS 


--The DAG


--Regression model in LaTeX


Interpretation and additional statistics


---
## The Assumptions

-- More complex DAG

--Assumptions
 1. No Treatment-Outcome Confounding
 2. No Mediator-Outcome Confounding
 3. No Treatment-Mediator Confounding
 4. No Mediator-Outcome confounding that is an effect of treatment (MSM)
 5. Temporal Ordering


---
class: inverse, middle, center
background-size: contain

# What is mediation analyses

> _"The presence or absence of statistical interaction between two factors depends on the scale with which one chooses to measure their association" -Modern Epidemiology_




---
## Mediation Assessment and Heuristics

-- Good tips and tricks you need to think about
-- This is specific to mediation but could have broader applications



---
## A tale of two dags 

two dag pics 

Should I Adjust for Z or no? 
Even if you or your investigator is not interested in Mediation you could be adjusting for a variable that is on the pathway and therefore creating a biased estimate of your effect







---
## Some good rules of thumb

muddled variables DISTINCT CONSTRUCTS 

muddled associations (oncology example)

Although design is not crucial it helps

See a lot of mediation analyses done in cross-sectional data 

Words of advice, mediation is worse for fishing, so most of your work you will need to do is not 

musings on causal estimation vs causal discovery

How to choose between interaction and mediation



--
## Data example 



```{r}
put awesome data here 
```




