---
title: "Mediation"
author: "Patrick Wilson MPH"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
solarized_dark(
  code_font_family = "Fira Code",
  code_font_url    = "https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css"
)
```


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE)

library(arsenal)
library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
```

## What we hope to go over

--

- Background 

--

- Causal Estimands and DAGS

--

- Mediation Assessment and Heuristics


--

- Example of Mediation Analyses

--

- Resources

---
class: inverse, middle, center
background-size: contain

#What is mediation analyses

> _"The presence or absence of statistical interaction between two factors depends on the scale with which one chooses to measure their association" -Modern Epidemiology_


---
##Overview 


---
##What does Causal inference add?


---
##What kind of questions can you ask?

- Pain Example

- Depression Example

- Biomarker Example

- Reproductive Epidemiology Example

- Social Epidemiology Example



- Let us use a classic example in observational epidemiology to illustrate the issue of scale

- Here you have a hypothetical cohort looking at the interaction between smoking and asbestos on the effect of getting lung cancer per 1,000 persons.


```{r, echo = FALSE}
data_2b2 <- data.frame( status = c("Non-smoker", "Smoker"), `No Asbestos` = c("$p_{00}$ = 1", "$p_{01}$ = 10"), `Asbestos` = c("$p_{10}$ = 3", "$p_{11}$ = 20")) %>% 
  rename(` ` = status)

kable(data_2b2, "html", escape =F) %>% 
  kable_styling("striped", full_width = F, font_size = 20) 
```

- With these four risks you can calculate everything you need to assess interaction.




